Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-08-26 inst/manuscript/manuscript.Rmd
some updates

@ver2024marginal show that the fixed effect estimator is $\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{w}$, where $\hat{\boldsymbol{\Sigma}}$ is the covariance matrix $\boldsymbol{\Sigma}$ evaluated at $\hat{\boldsymbol{\theta}}$.

@@

@ver2024marginal also derive the covariance matrix of $\hat{\boldsymbol{\beta}}$ and show how to predict $\mathbf{w}$ at unobserved locations and quantify their uncertainties.

>>>>

Though the fixed effects are integrated out of the likelihood, we can still estimate them using generalized least squares (GLS) principles, a common practice for linear models estimated using restricted maximum likelihood methods. Had we observed $\mathbf{w}$, a GLS estimator for $\boldsymbol{\beta}$ is given by
\begin{equation}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{equation}
where $\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}$. However, we only observe $\hat{\mathbf{w}}$, so it is reasonable to define $\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}$. Thus, to derive properties of $\hat{\boldsymbol{\beta}}$ like expectation and variance, we must derive these properties for $\hat{\mathbf{w}}$. To do so, we must condition on $\mathbf{w}$ as if it were observed and invoke properties of the laws of total expectation and variance. Because $\hat{\mathbf{w}}$ was optimized via the likelihood, we assume that given $\mathbf{w}$, $\hat{\mathbf{w}}$ has mean $\mathbf{w}$ and variance approximately equal to $-\mathbf{H}^{-1}$ (the inverse Hessian). It follows that $\text{E}(\hat{\mathbf{w}})$ is given by
\begin{equation}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{equation}
and $\text{Var}(\hat{\mathbf{w}})$ is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align}
Putting this all together, it follows that
\begin{equation}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}
\end{equation}
and
\begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
\end{align}
In practice, $\text{Var}(\hat{\boldsymbol{\beta}})$ is estimated by evaluating $\boldsymbol{\Sigma}$ at $\hat{\boldsymbol{\theta}}$, the estimated covariance parameter vector.

These results are important because they justify closed-form solutions for $\hat{\boldsymbol{\beta}}$ and its associated variance. Closed-form solutions are useful because they bypass the need for computationally expensive sampling-based strategies to evaluate the mean and variance of $\hat{\boldsymbol{\beta}}$ -- a common technique for other approaches to SPGLMs like Bayesian MCMC.

@@

We may also predict values of the latent mean (on the link scale) at new locations by leveraging the spatial covariance between observed locations and new locations. Again suppose that we observed $\mathbf{w}$ and we want to make predictions at $\mathbf{u}$, a vector of latent means at the new locations that follows the same SPGLM from Equation~\ref{eq-spglm} with fixed effects design matrix, $\mathbf{X}_{\mathbf{u}}$. The vector $(\mathbf{w}, \mathbf{u})^\top$ has the following properties:
\begin{align}
  \text{E}(\mathbf{w}, \mathbf{u})^\top & = (\text{E}(\mathbf{w}), \text{E}(\mathbf{u}))^\top = (\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top \\
  \text{Var}(\mathbf{w}, \mathbf{u})^\top & = \begin{bmatrix} \text{Var}(\mathbf{w}, \mathbf{w}) & \text{Var}(\mathbf{w}, \mathbf{u}) \\ \text{Var}(\mathbf{u}, \mathbf{w}) & \text{Var}(\mathbf{u}, \mathbf{u}) \end{bmatrix} = \begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}
\end{align}
Because we have observed $\mathbf{w}$, we may derive the conditional distribution of $\mathbf{u}|\mathbf{w}$, which has the following properties:
\begin{align}
  \text{E}(\mathbf{w} | \mathbf{u}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
  \text{E}(\mathbf{w} | \mathbf{u}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
\end{align}
Two more sources of uncertainty, $\hat{\boldsymbol{\beta}}$ and $\hat{\mathbf{w}}$.
Spatial prediction is often also called Kriging [@cressie1990origins].

### 2025-08-26 inst/manuscript/manuscript.tex
some updates

\citet{ver2024marginal} show that the fixed effect estimator is
\(\hat{\boldsymbol{\beta}} = (\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{X})^{-1}\mathbf{X}^T \hat{\boldsymbol{\Sigma}}^{-1}\mathbf{w}\),
where \(\hat{\boldsymbol{\Sigma}}\) is the covariance matrix
\(\boldsymbol{\Sigma}\) evaluated at \(\hat{\boldsymbol{\theta}}\).

@@

\citet{ver2024marginal} also derive the covariance matrix of
\(\hat{\boldsymbol{\beta}}\) and show how to predict \(\mathbf{w}\) at
unobserved locations and quantify their uncertainties.

>>>>

Though the fixed effects are integrated out of the likelihood, we can
still estimate them using generalized least squares (GLS) principles, a
common practice for linear models estimated using restricted maximum
likelihood methods. Had we observed \(\mathbf{w}\), a GLS estimator for
\(\boldsymbol{\beta}\) is given by \begin{equation}\label{eq-gls1}
  \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{w} = \mathbf{B}\mathbf{w},
\end{equation} where
\(\mathbf{B} = (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\).
However, we only observe \(\hat{\mathbf{w}}\), so it is reasonable to
define \(\hat{\boldsymbol{\beta}} = \mathbf{B}\hat{\mathbf{w}}\). Thus,
to derive properties of \(\hat{\boldsymbol{\beta}}\) like expectation
and variance, we must derive these properties for \(\hat{\mathbf{w}}\).
To do so, we must condition on \(\mathbf{w}\) as if it were observed and
invoke properties of the laws of total expectation and variance. Because
\(\hat{\mathbf{w}}\) was optimized via the likelihood, we assume that
given \(\mathbf{w}\), \(\hat{\mathbf{w}}\) has mean \(\mathbf{w}\) and
variance approximately equal to \(-\mathbf{H}^{-1}\) (the inverse
Hessian). It follows that \(\text{E}(\hat{\mathbf{w}})\) is given by
\begin{equation}
  \text{E}(\hat{\mathbf{w}}) = \text{E}(\text{E}(\hat{\mathbf{w}} | \mathbf{w}))
   = \text{E}(\mathbf{w}) 
   = \mathbf{X}\boldsymbol{\beta}
\end{equation} and \(\text{Var}(\hat{\mathbf{w}})\) is given by
\begin{align}
  \text{Var}(\hat{\mathbf{w}}) & = \text{E}(\text{Var}(\hat{\mathbf{w}} | \mathbf{w})) + \text{Var}(\text{E}(\hat{\mathbf{w}} | \mathbf{w})) \\
  & = \text{E}(-\mathbf{H}^{-1}) + \text{Var}(\mathbf{w})\\
  & = -\mathbf{H}^{-1} + \boldsymbol{\Sigma}
\end{align} Putting this all together, it follows that \begin{equation}
  \text{E}(\hat{\boldsymbol{\beta}})  = \text{E}(\mathbf{B}\hat{\mathbf{w}}) 
  = \mathbf{B}\text{E}(\hat{\mathbf{w}}) =  (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}(\mathbf{X}^\top \boldsymbol{\Sigma}^{-1} \mathbf{X})\boldsymbol{\beta} = \boldsymbol{\beta}
\end{equation} and \begin{align}
  \text{Var}(\hat{\boldsymbol{\beta}}) & = \text{Var}(\mathbf{B}\hat{\mathbf{w}}) \\
  & = \mathbf{B} \text{Var}(\hat{\mathbf{w}}) \mathbf{B}^\top\\
  & = \mathbf{B} (-\mathbf{H}^{-1} + \boldsymbol{\Sigma}) \mathbf{B}^\top \\
  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + \mathbf{B}\boldsymbol{\Sigma}\mathbf{B}^\top \\
  & = \mathbf{B}-\mathbf{H}^{-1}\mathbf{B}^\top + (\mathbf{X}^\top \boldsymbol{\Sigma}^{-1}\mathbf{X})^{-1}
\end{align} In practice, \(\text{Var}(\hat{\boldsymbol{\beta}})\) is
estimated by evaluating \(\boldsymbol{\Sigma}\) at
\(\hat{\boldsymbol{\theta}}\), the estimated covariance parameter
vector.

These results are important because they justify closed-form solutions
for \(\hat{\boldsymbol{\beta}}\) and its associated variance.
Closed-form solutions are useful because they bypass the need for
computationally expensive sampling-based strategies to evaluate the mean
and variance of \(\hat{\boldsymbol{\beta}}\) -- a common technique for
other approaches to SPGLMs like Bayesian MCMC.

@@

We may also predict values of the latent mean (on the link scale) at new
locations by leveraging the spatial covariance between observed
locations and new locations. Again suppose that we observed
\(\mathbf{w}\) and we want to make predictions at \(\mathbf{u}\), a
vector of latent means at the new locations that follows the same SPGLM
from Equation\textasciitilde{}\ref{eq-spglm} with fixed effects design
matrix, \(\mathbf{X}_{\mathbf{u}}\). The vector
\((\mathbf{w}, \mathbf{u})^\top\) has the following properties:
\begin{align}
  \text{E}(\mathbf{w}, \mathbf{u})^\top & = (\text{E}(\mathbf{w}), \text{E}(\mathbf{u}))^\top = (\mathbf{X}\boldsymbol{\beta}, \mathbf{X}_\mathbf{u}\boldsymbol{\beta})^\top \\
  \text{Var}(\mathbf{w}, \mathbf{u})^\top & = \begin{bmatrix} \text{Var}(\mathbf{w}, \mathbf{w}) & \text{Var}(\mathbf{w}, \mathbf{u}) \\ \text{Var}(\mathbf{u}, \mathbf{w}) & \text{Var}(\mathbf{u}, \mathbf{u}) \end{bmatrix} = \begin{bmatrix} \boldsymbol{\Sigma} & \boldsymbol{\Sigma}_{\mathbf{w}\mathbf{u}} \\ \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{w}} & \boldsymbol{\Sigma}_{\mathbf{u}\mathbf{u}} \end{bmatrix}
\end{align} Because we have observed \(\mathbf{w}\), we may derive the
conditional distribution of \(\mathbf{u}|\mathbf{w}\), which has the
following properties: \begin{align}
  \text{E}(\mathbf{w} | \mathbf{u}) & = \mathbf{X}_{\mathbf{u}} \boldsymbol{\beta} + \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}}\boldsymbol{\Sigma}^{-1}(\mathbf{w} - \mathbf{X}\boldsymbol{\beta}) \\
  \text{E}(\mathbf{w} | \mathbf{u}) & = \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{u}} - \boldsymbol{\Sigma}_{\mathbf{u}, \mathbf{w}} \boldsymbol{\Sigma}^{-1} \boldsymbol{\Sigma}_{\mathbf{w}, \mathbf{u}}
\end{align} Two more sources of uncertainty,
\(\hat{\boldsymbol{\beta}}\) and \(\hat{\mathbf{w}}\). Spatial
prediction is often also called Kriging \citep{cressie1990origins}.
